Efficiency

### 2018-01-07 DraftExec.Rnw
something's not working again.  Just get "missing $ inserted." no line ref.  Aargh

In stark contrast, a submarine operator might have less than 1\% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially \textbf{more} concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
%
This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.  Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.  There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load. It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.  (They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)

@@

Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.  Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.  Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available \parencite{WNA2017}.

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.  It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their \textbf{peak} output.

@@

Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.

@@

@@

Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.  It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.

@@

In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull. This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed. Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.

\subsubsection{Venturi Effect}
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is a simple embodiment of the Venturi effect.

@@

@@

\subsection{Modern literature and results}
In addition, more modern works, including those which involve experimental validation of theoretical results also tend to show that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.

@@

\subsection{Theoretical Limitations}
In this section I will appraise some of the available literature, including some of the most canonical works in the design of pumpjets which is now declassified from US Navy funded research (and still extremely frequently cited by contemporary researchers).  Before doing so, I'll attempt to summarise in simple language the mose essential conclusions in simple language.

My broad conclusion is that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.  These two design requirements can be found either to be in tension, or alignment, in different regimes.  At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.  Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.  However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.  In a variety of intermediate regimes, the two requirements find themselves in tension.

@@

The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.  In later sections I will address further the acoustic comparisons in the speed ranges at 10kt and below.

@@

Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.  If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards. Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large.

@@

It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, for a particular fixed vessel size and drag, and fixed propulsion system design smaller values of $\Delta V / V_0$ will correspond to to the propulsion system exerting less thrust, and also moving at a lower speed under self-propulsion, since the thust required (proportional to $\Delta V$ when operating at steady state) will decline faster than the vessel speed or water speed at intake, since the drag on the vessel scales with the square of the vessel speed.

@@

The structure of these overall curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.  Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}. It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.

@@

The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.  This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in Section \ref{Speed and Drag - Why very slow is very very $(very)^2$ economical}. Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two \parencite{greenfield2016}.

@@

Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.  For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{article{buckingham2008submarine} at around 4kt.  Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.  The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.

@@

As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.  It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{@stanford2017}.  Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.

@@

>>>>

In stark contrast, a submarine operator might have less than 1\% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially \textbf{more} concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.

This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.  Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.  There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load. It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.  (They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)

@@

\subsection{Power and Energy}
Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.  Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.  Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available \parencite{WNA2017}.

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.  It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their peak output.

@@

\subsection{Xenon poisoning and low power limitations}
Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.

@@

\subsection{Recent Commentary}
Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet, but on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

The authors of this piece also suggest that the high cost and weight of the nuclear reactor contributes to the desire to minimise hotel load.  Given the way that the square law drives drag as discussed earlier, driving a cube law for power, there is no chance at all that the hotel load has any bearing on the reactor power required.  The last few knots of top speed would demand ten times over the power required for hotel load, as can be seen in Figure \ref{fig:power_curve}.

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.  One would think that the sorts of emergencies or accidents which would drive this are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably in a couple of days.

This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.  It might be presumed that surfacing and remaining surfaced for a couple of days is likely to end combat operations for a submarine that was located in theatre.  However, remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted, especially since hotel load, in the case of the submarine they propose, amounts to more than propulsion at likely patrol speeds.  As we can gather from other public estimates of submarine submerged endurance \parencite{buckingham2008submarine}, roughly a couple of days is a very significant fraction of the entire dived endurance of a diesel-electric submarine, perhaps 50\%.  It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of diesel-electric submarines, in addition to the reactor.

Consequently, it seems much more likely that the sort of battery-requirements that are necessary for this kind of safety system are not closely linked to a full operational hotel load, but some substantially reduced survival load.  As discussed in later sections, many of the most energy intensive systems in the hotel load that could readily be switched off in an emergency to save power, if saving power was what was required to get through the emergency, including the combat system.  It is difficult to imagine what sort of emergency might require a reactor to be shut down, that wouldn't also justify shutting down essentially every other component of normal hotel load which could plausibly be delayed for a couple of days.

If it was known that the reactor couldn't be restarted after the poison-out period, the mission need to be aborted in any case, and some kind of rescue sought. Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine literally at the entrance of a home port.  It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.

@@

Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.  It says that the total energy in a connected body of fluid is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.

\begin{equation}
\label{eq:bernoullis}
\frac{1}{2}\rho v^2 + \rho gz + p = C
\end{equation}

@@

In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull. This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed. Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.  In the case of submarine pumpjets, located on the end of the hull and parallel to the axis of the submarine, this effect (and the term $gz$) are not relevant.

\subsubsection{Venturi Effect}
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is a simple embodiment of the Venturi effect.

@@

Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.  Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.  This is the reason that golf balls have dimples, and tennis-balls have fur.

@@

\subsection{Modern literature and results}
In addition, more modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.  It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.  Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.  However, from a few notable studies exist which provide important reference points. (Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)

@@

\subsection{A simple theoretical explanation}

Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.  They don't just accelerate a water colum.  They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.  When the total amount of energy that the incoming water has, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.  The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.  In essence, if you try to slow down something that's already going slowly, it tends to stop.  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, takes place.

There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.  They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.  the the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.

The conditions which lead to flow separation in a pumpjet can be relatively simply explained in terms of conservation of energy, and conservation of mass. Conservation of mass requires that as a pipe or duct expands, the flow through it necessarily slows down, in order to occupy the large volume, in proportion to the new cross-sectional area.  Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.  That relationship says that the static pressure increase is inversely proportional change in velocity **squared**, as given by Bernoulli's equation, which was discussed in earlier sections.  As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.

It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.  The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.  At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.  This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship.

At low speeds, the same effect leads to detrimental effects.  The adverse pressure gradient is much reduced, but the velocity is much much $(much)^2$ reduced. The pressure gradient is necessarily constant across the entire area radial area of the duct, but the drag due to the boundary with the duct itself will necessarily be greater closer to the walls in the case of the intake of the duct.  For this simple reason, as the flow rate decreases, some point is reached where the flow adjacent to the walls actually reverses, and the flow becomes separated.

The same phenomenon is also prone to occur near roots of the blades of the rotor.  Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase through the blades, providing an adverse pressure gradient, which again must be constant radially.  Since closer to the hub the velocity of the blads through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid. With the longer surface over which the fluid must travel, as well as greater angle of deflection, cavitation is prone to occur here too.  The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different. This will also mean that the direction of rotation of the vortices will be different, and the location also close to the middle of the jet, instead of the outside.

As can be seen in figure \ref{fig:JetHighFlow.png}, the duct of a jet is design to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.   Instead, pumpjets tend to use some moderate expansion of the flow in advance of the impeller (diffusion) in order to raise its pressure, and reduce its velocity.  (Allowing the blades to do work on the water at lower speeds and higher pressure is the essence of what reduces cavitation.)  The blades then do their work on the liquid, further raising its pressure.  The duct allows this pressure then to be converted back to kinetic energy, as the water accelerates to move through the contracting area in the nozzle.  This difference in the shape of the flow, and the consequent impacts on the static pressures, and demonstrated in Figure \ref{fig:FlowComparison.png} taken from \cite{mccormick1963design}.

\begin{figure}
\includegraphics[width=\textwidth]{FlowComparison.png}
\caption{The effect of the duct is to deflect the flow away from the streamlines of an open propeller, elevating the static pressure at the stator blades by slowing the flow down.}
\label{fig:FlowComparison.png}
\end{figure}

Consequently, the geometry of the duct defines a pressure gradient in the intake and through the nozzle.  The greater the difference in the cross-sectional areas, the greater the total magintude of the pressure differential that will occur, as necessitated by conservation of mass. The same amount must go out as came in.  As the area through which it passes changes, the velocity much change to that the flow rate remains constant, otherwise cavitation will occur.  The rate at which the areas change (how rapidly they expand and contract along the axis) will determine the gradient.  Making the pressure change as smooth as possible is an important aspect to jet design in order to minimise flow separation.

\begin{figure}
\includegraphics[width=\textwidth]{JetHighFlow.png}
\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red. After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
\label{fig:JetHighFlow.png}
\end{figure}

Figure \ref{fig:JetLowFlow.png} shows a jet operating at very low speed.  The adverse pressure gradient combined with the drag forces of the boundary layer are sufficient to bring the flow to a stop and reverse it, as shown by the arrow size reducing, and changing direction in the diagram.  This occurs very close to the inner surface near the opening of the duct, and potentially also between the blades of the rotor, beginning the roots \parencite[15]{henderson1964}.  Flow separation necessarily sees additional energy expended in the kinetic energy of the additional vortices and turbulent flows created, which don't result in the creation of thrust, which demands a change in the net momentum of the liquid in a backwards direction.  Consequently, the degree and extent that it occurs necessarily and inevitably reduces the efficiency of the propulsor.  As the flow rates slow further, the areas of flow separation will become larger, effectively reducing the intake area of the jet, and further degrading efficiency.

Flow separation also tend to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.  As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.  These effects also have acoustic consequences, which will be discussed later.

\begin{figure}
\includegraphics[width=\textwidth]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}

It is worthwhile noting that the diagrams in Figures \ref{fig:JetHighFlow.png} and \ref{fig:JetLowFlow.png} do not include a hub for the sake of simplicity.  In practice, the cross-sectional are of the pumpjet naturally only includes that aread between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.  Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier argued it generally contracts.  It is important to consider the impact of the second dimension on the actual effective area. Since area of a disc follows $\pi r^2$, the impact of the contraction of the hub will be dramatically larger when it is large, than it will be when it is small.

Consequently, the contraction of the outer shroud at the very end of the duct will dominate the final taper of the hub, and actually lead to substantial restriction of the flow.  Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area available, and considerable diffusion.  The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practially realistics to avoid flow separation, even at quite high speeds. It is important to recognise that this simple relationship with $r^2$ generally means that even moderate changes in terms of diameter, which is what we tend to observe when looking at images and diagrams can actually amount to much more substantial changes in velocity, and hence pressure, which drives the effects described in this section.  A side-by-side comparison of the geometries and flow-lines likely in the case of pumpjet or propeller is give in Figure \ref{fig:Comparison.png}, from  to help illustrate this point.

\begin{figure}
\includegraphics[width=\textwidth]{Comparison.png}
\caption{At the end of a tapering body the impact of the contracting hub will be much larger near the intake than near the tail, making the degree of diffusion ahead of the impeller, as well as contraction of the flow in the nozzle after it, much larger in impact than it appears by looking at the external shape. Credit \cite{gearhart1966selection}}
\label{fig:Comparison.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{Triomph.jpg}
\caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves. Credit \cite{reddit2016}}
\label{fig:Triomph.jpg}
\end{figure}

\subsection{Trade-offs and Limitations on Improvements}

Having established that the inevitable decline of a pumpjets efficiency towards lower speeds is driven by the increasing occurrence of flow separation, some consideration ought to be given to the potential scope for the effect to be improved or negated by design changes.  Here I have found considerable help from well established literature, many including various old declassified technical reports funded by the US Navy, many of which are frequently cited in the most recent literature as laying out the enduring frameworks for the modern discipline.  My reading on the matter which reminds me strongly of an adage recently quoted in Senate Estimates by General Manager for Submarines, Stephen Johnson: "There's no free lunch in physics."

It is clear that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.  These two design requirements can be found either to be in tension, or alignment, in different regimes.  At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.  Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.  However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.  In a variety of regimes (including at quite high speeds), the two requirements find themselves in tension and needing to be balanced \parencite{gearhart1966selection}.

@@

The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.  In my research I have found no evidence of waterjets being advocated for speeds less than 30kt, and for pumpjets for speeds less than 27kt.  Having consulted with leading academics in the field, I've had it strongly confirmed that a pumpjet will always be less efficient than a propeller at speeds under 18kt.  In later sections I will address further the acoustic comparisons at low speeds where flow separation will begin to occur, for instance below 10kt.

@@

Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.  If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards. Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large, as discussed earlier.

@@

It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this value diminishing towards zero at very low speeds.  In an idealised circumstance, the flow rate increases directly with vessel speed.  Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.  However, in practice the water encountered by the propulsor is slowed by the wake field, the impact of which is likely non-linear with vessel speed.  In essence, equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.  This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.  This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.  Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.

@@

The structure of these curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.  Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}. It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.

@@

The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.  This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in earlier sections. Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two \parencite{greenfield2016}.

@@

Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.  For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.  Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.  The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.

@@

As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.  It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.  Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.

@@

The precise point at which cavitation inception occurs, what type of cavitation it is, and where it occurs on the blades is a complex science, with considerable work being continually undertaken to validate theoretical models. A good overview of the some of the science can be found in reference \cite{kuiper1981cavitation}.  Still today, similar techniques such as paint tests and visual observations are validate models and theory. However, despite the inherent complexity about predicting the precise point in time and space of onset, and the respective impact on efficiency and acoustics, several broad relationships are known to hold well, and can be relied upon to characterise the most significant trends that determine when cavitation will be more likely to occur, which is generally driven by the diffence between the vapour pressure of water and the local static pressure.
